An enhancer code compendium of early human neural and neural crest development

### 2025-04-14 manuscript.tex
changes S.A.; changes C.M.

\title{An enhancer code compendium of early human neuronal and neural crest development}
\date{}

@@

\author[1,2,3,\orcidlink{0000-0001-7907-1247}]{Seppe De Winter}
\author[1,2,3, 4, \orcidlink{0000-0002-1318-2603}]{Camiel Mannens}
\author[1,2,3]{Valerie Christiaens}
\author[4, \orcidlink{0000-0003-1869-0372}]{Lijuan Hu}
\author[5, \orcidlink{0000-0003-2931-8015}]{Erik Sundström}
\author[4, \orcidlink{0000-0002-3491-3444}]{Sten Linnarsson}
\author[1,2,3,*,\orcidlink{0000-0002-8006-0315}]{Stein Aerts}

@@

\affil[1]{VIB Center of AI \& computational Biology, Leuven, Belgium}
\affil[2]{VIB-KU Leuven Center for Brain and Disease Research, Leuven, Belgium}
\affil[3]{KU Leuven Department of Human Genetics, Leuven, Belgium}
\affil[4]{Karolinska Institute, Department of Biochemistry and Biophysics, Stockholm, Sweden}
\affil[5]{Karolinska Institute, Division of Neurodegeneration, Department of Neurobiology, Care Sciences and Society,Stockholm, Sweden}

@@

%Context
Enhancers encode cell types as clusters of transcription factor binding sites. 
%need
Enhancer code studies are predominantly limited to a small number of cell types often focusing
on model organisms.
%task and results
In this study we generate a single-cell multiome atlas of human neural tube organoids 
and of the head of a four weeks post conception human embryo. 
Using DNA sequence based deep learning - DeepNeuralTube - we identify a total of 15 cell types that 
are shared between the organoid model and the human embryo and we provide a compendium of enhancer codes 
for each individual cell type. 
We confirm the paradigm that enhancers are encoded using heterotypic clusters of 4-5 transcription factor binding sites in a nucleosome depleted region. 
Enhancers encoding anterior-posterior patterning form an exception to this rule where they are encoded using
homotypic clusters of homeodomain binding sites.
%conclusion
This compendium of enhancer codes thus provides gene regulatory insight into human development.
%implication

@@

Genomes of complex multi-cellular organisms specify all cell types and states\cite{trapnellDefiningCellTypes2015, arendtOriginEvolutionCell2016, fleckWhatCellType2023} through distinct non-coding elements named
enhancers\cite{davidsonRegulatoryGenomeGene2010}. 
Central to the study of the genomic code is the paradigm that 
enhancers consist of heterotypic clusters of transcription factor binding sites (TFBS)\cite{longEverChangingLandscapesTranscriptional2016,kimDecipheringMultiscaleQuantitative2023} with soft syntactic rules\cite{zeitlingerSevenMythsHow2020} 
and that cooperative binding of multiple transcription factors (TFs)\cite{lambertHumanTranscriptionFactors2018} underlies cell type-specific activity of the enhancer.\par

DNA sequence based deep learning methods\cite{
  alipanahiPredictingSequenceSpecificities2015,
  zhouPredictingEffectsNoncoding2015,
  kelleyBassetLearningRegulatory2016,
  quangDanQHybridConvolutional2016,
  movvaDecipheringRegulatoryDNA2019,
  kelleySequentialRegulatoryActivity2018,
  yuanScBassetSequencebasedModeling2022,
  chenSequencebasedGlobalMap2022,
  maslovaDeepLearningImmune2020,
  minnoyeCrossspeciesAnalysisEnhancer2020,
  avsecBaseresolutionModelsTranscriptionfactor2021,
  avsecEffectiveGeneExpression2021,
  dealmeidaDeepSTARRPredictsEnhancer2022,
  linderPredictingRNAseqCoverage2025a,
  kempynckCREstedModelingGenomic2025} have shown major advances in the ability to model and explain enhancers\cite{zeitlingerSevenMythsHow2020,dewinterModellingDesignTranscriptional2025} 
and largely support the enhancer code paradigm. 
However, these studies are predominantly limited to a small number of cell type or cell states\cite{
  maslovaDeepLearningImmune2020,
  minnoyeCrossspeciesAnalysisEnhancer2020,
  atakInterpretationAllelespecificChromatin2021,
  avsecEffectiveGeneExpression2021,
  bravogonzalez-blasSinglecellSpatialMultiomics2024}, and often focus on model organisms\cite{
  janssensDecodingGeneRegulation2022,
  heckerEnhancerdrivenCellType2025}, with only few studies profiling human development\cite{mannensChromatinAccessibilityHuman2024}.\par

@@

In this study, we address these limitations by building an enhancer code compendium of cell types and cell states generated by human neural tube organoids, 
a model system for early human neuronal and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021}.
We compare this compendium to that of the head of a four weeks post conception human embryo.
Through these comparisons we confirm the enhancer code paradigm, 
but also provide a counter examples specifically related to the enhancer code underlying anterior-posterior patterning.
\par

@@

\subsection*{Human neural tube organoids recapitulate early neuronal and neural crest development}

We generated human neural tube organoids to model early human neuronal and neural crest development. 
To this end we embedded single cells from two induced pluripotent stem cell (iPSC) lines (Sigma and BJ1) and one embryonic stem cell line (ESC; H9) into separate synthetic poly-ethylene glycol (PEG) based extracellular matrices. The organoids were incubated for 11 days and we supplemented the medium with retinoic acid (RA) and Smoothened agonist (SAG) from day 3 until day 5 (\textbf{Fig. 1a}). 
These conditions were previously shown to generate multiple neuronal progenitor and
neural crest states\cite{abdelfattahActuationEnhancesPatterning2021}. 
To characterize the organoid-derived cell types to a greater level of detail at both transcriptomic as well as chromatin accessibility level we performed 10x single-cell multiomics (combined scATAC-seq and
scRNA-seq) 
at four time points of differentiation (day 4-5, day 6-7, day 8-9 and day 10-11; \textbf{Fig. 1a}), 
resulting in 22,862 high quality cells (\textbf{Fig. 1b}). 
Note, that similar transcriptomic states were obtained from all three cell lines (\textbf{Supplementary Fig. S1a-b}) demonstrating the robustness of the organoid system.\par

To evaluate the biological relevance of the cell types and states obtained from the organoid, 
we also performed 10x single cell multiome on the head of a single four weeks post conception human embryo, 
resulting in 25,785 high quality cells (\textbf{Fig. 1c}). 
We annotated both the organoid and embryo cells based on the expression of marker genes
(\textbf{Supplementary Table S1}). 
In both systems we could identify neuronal progenitors, early differentiating neurons,
neural crest, facial mesenchyme and early differentiating peripheral neurons (\textbf{Fig. 1b-c}).
Additionally, in the organoid only we could identify a pluripotent stem cell population (\textbf{Fig. 1b}) 
while perivascular macrophages (PVM) / microglia (MIC) and otic placode cells were missing (\textbf{Fig. 1c}).\par

Next, we sub-clustered the neuronal progenitor, early neuron, and neural crest cells (\textbf{Fig. 1d-f}). 
Both organoid and embryo neuronal progenitors show a gradient of different cell states reflecting dorso-ventral patterning, evident from the expression of dorso-ventral marker TFs\cite{delasRepressiveInteractionsGene2020}:
\textit{ARX}, \textit{NKX2-2}, \textit{PAX6}, \textit{IRX3}, \textit{ZIC1} and \textit{GDF7} (\textbf{Fig. 1d}). 
In both the organoid and embryo system we obtained multiple early neuronal differentiation states.
This includes a cluster that expresses the neuronal progenitor marker \textit{SOX2}, cells that express the early neuronal induction marker \textit{ASCL1} and distinct clusters marked by the expression of \textit{GAD1}, \textit{UNCX} and \textit{GATA2} (\textbf{Fig. 1e}). 
Finally, focusing on the neural crest, in both the organoid and the embryo we identified a population of pre-migratory neural crest (\textit{GRHL2} expression and absence of \textit{ZEB2} expression), migratory neural crest (\textit{SOX10} and \textit{ZEB2} expression), facial mesenchyme (\textit{TWIST1} expression) and early differentiation peripheral neurons (\textit{NEUROD1} expression). 
In the embryo, but not in the organoid, we observed an additional population of cells marked by the expression of \textit{SOX10} and \textit{PAX2} reflecting the otic placode (\textbf{Fig. 1f}).\par

@@

Scoring cells of both organoid and embryo origin for the expression of cell cylce marker genes, 
we observed that both early differentiating peripheral neurons and to a greater extent early differentiating neurons of the central nervous system are in cell cycle arrest (\textbf{Fig. 1h}), an important hallmark of neuronal differentiating\cite{
braunComprehensiveCellAtlas2023}, while other cell types are still in different phases of the cell cycle (\textbf{Fig. 1h})\par

Finally, to assess whether the organoid-derived cell types and states are also similar to the embryo on a chromatin accessibility level 
we identified sets of co-accessible regions in the organoid model using topic modeling 
and identified equivalent cells in the embryo where these same regions were also co-accessible.
We observed similar patterns of marker gene expression for the same cell state across the organoid and embryo (\textbf{Fig. 1i}) 
with an average Pearson correlation coefficient of 0.83 across identical states (\textbf{Supplementary Fig. S1c-d}).\par

@@

\afterpage{\clearpage}
\begin{figure}[p]
  \includegraphics[width=1.0\linewidth]{figures/main/Figure_1.png}
  \caption{
    \textbf{Human neural tube organoids recapitulate early neuronal and neural crest development.} \\
    \textbf{a}, Schematic of organoid differentiating protocol and human embryo experiment 
    \textbf{b-c}, UMAP based on gene expression profiles of organoid (\textbf{b}) and embryo
    (\textbf{c}) cells colored by cell type identify. 
    (Continues next page)
  }
\end{figure}
\afterpage{\clearpage}

\begin{figure}[b!]
  \ContinuedFloat
  \caption*{\rule{\linewidth}{0.4pt}}
  \caption{
    (Continued)
    \textbf{d-f}, UMAP based on gene expression profiles of organoid (top) and embryo (bottom) cells types showing sub-clusters of neuronal progenitors (\textbf{d}), early differentiating neurons (\textbf{e}) and neural crest (\textbf{f}) along with expression of indicated genes in red-green-blue color scale.
    \textbf{g}, UMAP based on gene expression profiles of organoid (top; UMAP axis 1 and 2 shown) and embryo (bottom; UMAP
    axis 2 and 3 shown) for neuronal progenitor subcluster along with the expression of indicated genes in red-green-blue
    color scale. 
    \textbf{h}, UMAP based on gene expression profiles in organoid (top left) and embryo (top right) colored by
    cell cycle phase identity and quantification of percentage of each cell cycle phase stratified by cell
    type in the organoid (bottom left) and embryo (bottom right). 
    \textbf{i}, Average gene expression level in $log_{2}(CPM)$ (dot color) and fraction of cells expressing the gene (dot size) per cell state, defined by topic modeling, for organoid
    cells (top) and embryo cells (middle) and motif enrichment level as normalized enrichment score (NES; dot color) and
    number of regions enriched for the motif (dot size) per topic (x-axis) and motif (y-axis; bottom). PVM/MIC: perivascular
    microphage/microglia.
    Color scales for RGB scatter plots are as follows (in Log(CPM)). 
    For organoid: 
    Minimum is 0 and maximum is 1.5 for all genes except:
    IRX3, BAHRL1: [0, 1.0], 
    FOXG1, GAD1, UNXC, GATA2, SOX10: [0, 0.8] and 
    ZEB2 [0, 3.0], 
    For embryo: 
    Minimum is 0 and maximum is 1.5 for all genes except:
    FOXG1, GAD1, UNCX, GATA2, SOX10 [0, 0.8], 
    ISL1 [0, 1.0], 
    ZEB2 [0, 3.0] and
    TWIST1 [0, 2.0].
  }
\end{figure}

\subsection*{
  DeepNeuralTube: sequence-to-function models that reveal the enhancer code underlying early neuronal and neural crest 
  development
}

In order to gain insight into the sequence characteristics underlying cell type-specific chromatin accessibility of 
neuronal progenitor, neural crest and neuronal cell type-specific enhancers, 
we trained two sequence-to-function model convolutional neural networks:
"$DeepNeuralTube^o$" and "$DeepNeuralTube^e$". 
The models use the DNA sequence of genomic regions as input and classify regions based on their cell type- or state-specific chromatin accessibility (in a multi-label multi-class fashion; \textbf{Fig. 2a}). 
The classes of the models are defined based on topic modeling. 
To get a fine-grained definition of cell states, we performed topic modeling separately for the subset of neuronal progenitors, neural crest (incl. facial mesenchyme and peripheral neurons) and early differentiating neurons for both organoid (to train $DeepNeuralTube^o$) embryo (to train $DeepNeuralTube^e$).\par

We tested the performance of $DeepNeuralTube^o$ and $DeepNeuralTube^e$ to predict cell type-specific activity 
using data from the VISTA enhancer browser\cite{viselVISTAEnhancerBrowsera2007, kosickiVISTAEnhancerBrowser2025}.
Enhancers active in the neural tube (n=349) could be classified with an area under the receiver-operator curve
(auROC) of 0.75 and 0.80 using resp. $DeepNeuralTube^o$ and $DeepNeuralTube^e$ (\textbf{Fig. 2b}; and
area under the precision-recall curve (auPR) of 0.21 resp. 0.29 (\textbf{Supplementary Fig. S2a-b})). 
Enhancers active in the facial mesenchyme (n=143) could be classified with an auROC of 0.78 and 0.79 
using resp. $DeepNeuralTube^o$ and $DeepNeuralTube^e$ (\textbf{Fig. 2c}; 
and auPR of 0.17 for both models (\textbf{Supplementary Fig. S2a-b})). 
Of note, a fraction of neural tube enhancers (\SI{24}{\percent}) are predicted to be specific to pre-migratory neural crest by
both models. Given that the VISTA enhancers are annotated based on localization of enhancer activity only, and given
that it is difficult to visually distinguish between the neural tube and pre-migratory neural crest, this result 
is not surprising.\par

Next, we assessed the explainability of the models. Concretely we tested if the model learned known
TF binding motifs as features. To this end, we calculated pairwise similarities\cite{
  guptaQuantifyingSimilarityMotifs2007} between \textit{de novo} motifs learned by the model, 
and identified using TF-MoDISco\cite{shrikumarTechnicalNoteTranscription2018}, and known motifs that are enriched in cell type specific regions.
Clustering this pairwise similarity matrix resulted in 56 clusters (\textbf{Fig. 2d}). 
All clusters, except cluster 47 (organoid specific CG motif), contain known motifs enriched in regions specific to both 
organoid and embryo cell types (\textbf{Fig. 2e} and \textbf{Supplementary Table S2}).
More importantly, Most clusters (\SI{70}{\percent}) contain at least one motif learned \textit{de novo} by the organoid and/or embryo model (\textbf{Fig. 2f-g} and \textbf{Supplementary Table S2}). 
Thus, $DeepNeuralTube^o$ and $DeepNeuralTube^e$ use known-motifs as features to classify regions based
on cell type-specific chromatin accessibility.\par

As an illustrative example, we investigated an enhancer located in one of the introns of \textit{GLI3} (hs111.0; \textbf{Supplementary Fig. S2c}). 
This region has neural tube specific activity in E11.5 mouse embryos (\textbf{Fig. 2h}). 
Both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ explain this region similarly (Pearson correlation of 0.79) 
and highlight the importance of multiple SOX TFBS (\textbf{Fig. 2h}). 
A version of this enhancer where 5\% of the nucleotides are changed (hs111.1) does not have neural tube specific enhancer activity and both models predict that this loss of activity is due to the loss of a sinlge SOX TFBS (\textbf{Fig. 2h}).\par

@@

\subsection*{
  The enhancer code underlying dorso-ventral patterning of neuronal progenitors
}

Focusing on neuronal progenitors, we identified topics representing cells and regions of different dorso-ventral
identities in both the organoid and the embryo data (\textbf{Fig. 3a}). 
From ventral to dorsal this is topic 36, 38, 33, 54 and 48 for the organoids and
topic 34, 38, 79, 88 and 58 for the embryo. 
As an example we first investigated two enhancers of \textit{SHH} that are strongly conserved and specific to the 
floorplate: SFPE1 and SFPE2\cite{jeongDistinctRegulatorsShh2003}. 
\textit{SHH} is most strongly expressed in the most ventral topic (33 and 34 resp. for the organoids and the embryo; \textbf{Fig. 3b}).
SFPE1 is located upstream of \textit{SHH} and SFPE2 in its second intron. Both enhancers are indeed most accessible in the cells belonging to the ventral topic (\textbf{Fig. 3c}). 
Explaining the accessibility of both enhancers using DeepNeuralTube reveals two FOX (one partial) and two RFX (one partial) binding sites in SFPE1 and two FOX, a ZEB and a TEAD binding site in SFPE2 (\textbf{Fig. 3d}).\par

>>>>

\title{An enhancer code compendium of early human neural and neural crest development}
\date{}

@@

\author[1,2,3,\orcidlink{0000-0001-7907-1247}]{Seppe De Winter}
\author[1,2,3, 4, \orcidlink{0000-0002-1318-2603}]{Camiel Mannens}
\author[1,2,3]{Valerie Christiaens}
\author[4, \orcidlink{0000-0003-1869-0372}]{Lijuan Hu}
\author[5, 6, 7, \orcidlink{0000-0002-2628-7834}]{Ruth M. Williams}
\author[5, 6, \orcidlink{0000-0001-9289-0263}]{Tatjana Sauka-Spengler}
\author[4, \orcidlink{0000-0002-3491-3444}]{Sten Linnarsson}
\author[1,2,3,*,\orcidlink{0000-0002-8006-0315}]{Stein Aerts}

@@

\affil[1]{VIB Center of AI \& computational Biology, Leuven, Belgium}
\affil[2]{VIB-KU Leuven Center for Brain and Disease Research, Leuven, Belgium}
\affil[3]{KU Leuven Department of Human Genetics, Leuven, Belgium}
\affil[4]{Karolinska Institute, Department of Biochemistry and Biophysics, Stockholm, Sweden}
\affil[5]{Stowers Institute for Medical Research, Kansas City, Missouri, United States of America}
\affil[6]{University of Oxford, MRC Weatherall Institute of Molecular Medicine, Radcliffe Department of Medicine, Oxford, United Kingdom}
\affil[7]{University of Manchester, Faculty of Biology, Medicine and Health, Manchester, United Kingdom}

@@

%Context
Enhancers encode cell types as clusters of transcription factor binding sites. 
%need
Enhancer code studies have been limited to specific cell types in model organisms.
%task and results
In this study we generate a single-cell multiome atlas of human neural tube organoids 
and of the head of a four weeks post conception human embryo. 
Using DNA sequence based deep learning - DeepNeuralTube - we identify 15 cell types that 
are shared between the organoid model and the human embryo and we provide a compendium of enhancer codes 
for each individual cell type. 
We confirm the paradigm that enhancers are encoded using heterotypic clusters of 4-5 transcription factor binding sites in a nucleosome depleted region. 
Enhancers encoding anterior-posterior patterning form an exception to this rule, being encoded as
homotypic clusters of homeodomain binding sites.
% TODO: S.A. [some more results? Validation examples? Genetic variants?]
%conclusion
This compendium of enhancer codes provides gene regulatory insight into human development.
%implication

@@

Genomes of complex multi-cellular organisms specify all cell types and states\cite{trapnellDefiningCellTypes2015, arendtOriginEvolutionCell2016, fleckWhatCellType2023} through distinct non-coding elements named
enhancers\cite{davidsonRegulatoryGenomeGene2010}. 
Central to the study of the genomic regulatory code is the paradigm that 
enhancers consist of heterotypic clusters of transcription factor binding sites (TFBS)\cite{longEverChangingLandscapesTranscriptional2016,kimDecipheringMultiscaleQuantitative2023} with soft syntactic rules\cite{zeitlingerSevenMythsHow2020}, 
and that cooperative binding of multiple transcription factors (TFs)\cite{lambertHumanTranscriptionFactors2018} underlies cell type-specific activity of the enhancer.\par

DNA sequence based deep learning methods\cite{
  alipanahiPredictingSequenceSpecificities2015,
  zhouPredictingEffectsNoncoding2015,
  kelleyBassetLearningRegulatory2016,
  quangDanQHybridConvolutional2016,
  movvaDecipheringRegulatoryDNA2019,
  kelleySequentialRegulatoryActivity2018,
  yuanScBassetSequencebasedModeling2022,
  chenSequencebasedGlobalMap2022,
  maslovaDeepLearningImmune2020,
  minnoyeCrossspeciesAnalysisEnhancer2020,
  avsecBaseresolutionModelsTranscriptionfactor2021,
  avsecEffectiveGeneExpression2021,
  dealmeidaDeepSTARRPredictsEnhancer2022,
  linderPredictingRNAseqCoverage2025a,
  kempynckCREstedModelingGenomic2025} have shown major advances in the ability to model and explain enhancers\cite{zeitlingerSevenMythsHow2020,dewinterModellingDesignTranscriptional2025} 
and largely support this enhancer code paradigm. 
However, these studies have been predominantly limited to a small number of cell types or cell states\cite{
  maslovaDeepLearningImmune2020,
  minnoyeCrossspeciesAnalysisEnhancer2020,
  atakInterpretationAllelespecificChromatin2021,
  avsecEffectiveGeneExpression2021,
  bravogonzalez-blasSinglecellSpatialMultiomics2024}, and often focus on model organisms\cite{
  janssensDecodingGeneRegulation2022,
  heckerEnhancerdrivenCellType2025}, with only few studies profiling human development\cite{mannensChromatinAccessibilityHuman2024}.\par

@@

In this study, we address these limitations by building an enhancer code compendium of cell types and cell states generated by human neural tube organoids, 
a model system for early human neural and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021}.
We compare this compendium to that of the head of a four weeks post conception human embryo.
Through these comparisons we assess the enhancer code paradigm, 
and also provide counter examples specifically related to the enhancer code underlying anterior-posterior patterning.
\par

@@

\subsection*{Human neural tube organoids recapitulate early neural and neural crest development}

We generated human neural tube organoids to model early human neural and neural crest development. 
To this end we embedded single cells from two induced pluripotent stem cell (iPSC) lines (Sigma and BJ1) and one embryonic stem cell line (ESC; H9) into separate synthetic poly-ethylene glycol (PEG) based extracellular matrices. The organoids were incubated for 11 days and we supplemented the medium with retinoic acid (RA) and Smoothened agonist (SAG) from day 3 until day 5 (\textbf{Fig. 1a}). 
These conditions were previously shown to generate multiple neural progenitor and
neural crest states\cite{abdelfattahActuationEnhancesPatterning2021}. 
To characterize the organoid-derived cell types to a greater level of detail at both transcriptomic as well as chromatin accessibility level we performed 10x single-cell multiomics (combined scATAC-seq and
scRNA-seq) 
at four time points of differentiation (day 4-5, day 6-7, day 8-9 and day 10-11; \textbf{Fig. 1a}), 
resulting in 22,862 high quality cells (\textbf{Fig. 1b}). 
Note that similar transcriptomic states were obtained from all three cell lines (\textbf{Supplementary Fig. S1a-b}) demonstrating the robustness of the organoid system.\par

To evaluate the biological relevance of the cell types and states obtained from the organoid, 
we also performed 10x single cell multiomics on the head of a single four weeks post conception human embryo, 
resulting in 25,785 high quality cells (\textbf{Fig. 1c}). 
We annotated both the organoid and embryo cells based on the expression of marker genes
(\textbf{Supplementary Table S1}). 
In both systems we could identify neural progenitors, early differentiating neurons,
neural crest, facial mesenchyme and early differentiating peripheral neurons (\textbf{Fig. 1b-c}).
Additionally, in the organoid only we could identify a pluripotent stem cell population (\textbf{Fig. 1b}) 
while perivascular macrophages (PVM) / microglia (MIC) and otic placode cells were missing in the organoid system (\textbf{Fig. 1c}).\par

Next, we sub-clustered the neural progenitor, early neuron, and neural crest cells (\textbf{Fig. 1d-f}). 
Both organoid and embryo neural progenitors show a gradient of different cell states reflecting dorso-ventral patterning, evident from the expression of dorso-ventral marker TFs\cite{delasRepressiveInteractionsGene2020}:
\textit{ARX}, \textit{NKX2-2}, \textit{PAX6}, \textit{IRX3}, \textit{ZIC1} and \textit{GDF7} (\textbf{Fig. 1d}). 
In both the organoid and embryo system we obtained multiple early neural differentiation states.
This includes a cluster that expresses the neural progenitor marker \textit{SOX2}, cells that express the early neural induction marker \textit{ASCL1} and distinct clusters marked by the expression of \textit{GAD1}, \textit{UNCX} and \textit{GATA2} (\textbf{Fig. 1e}). 
Finally, focusing on the neural crest, in both the organoid and the embryo we identified a population of pre-migratory neural crest (\textit{GRHL2} expression and absence of \textit{ZEB2} expression), migratory neural crest (\textit{SOX10} and \textit{ZEB2} expression), facial mesenchyme (\textit{TWIST1} expression) and early differentiation peripheral neurons (\textit{NEUROD1} expression). 
In the embryo, but not in the organoid, we observed an additional population of cells marked by the expression of \textit{SOX10} and \textit{PAX2} reflecting the otic placode (\textbf{Fig. 1f}).\par

@@

Scoring cells of both organoid and embryo origin for the expression of cell cylce marker genes, 
we observed that both early differentiating peripheral neurons and to a greater extent early differentiating neurons of the central nervous system are in cell cycle arrest (\textbf{Fig. 1h}), an important hallmark of neural differentiating\cite{
braunComprehensiveCellAtlas2023}, while other cell types are still in different phases of the cell cycle (\textbf{Fig. 1h})\par

Finally, to assess whether the organoid-derived cell types and states are also similar to the embryo at the level of chromatin accessibility
we identified sets of co-accessible regions in the organoid model using topic modeling 
and identified equivalent cells in the embryo where the same regions were also co-accessible.
We observed similar patterns of marker gene expression for the same cell state across the organoid and embryo (\textbf{Fig. 1i}) 
with an average Pearson correlation coefficient of 0.83 across matched states (\textbf{Supplementary Fig. S1c-d}).\par

@@

\afterpage{\clearpage}
\begin{figure}[p]
  \includegraphics[width=1.0\linewidth]{figures/main/Figure_1.png}
  \caption{
    \textbf{Human neural tube organoids recapitulate early neural and neural crest development.} \\
    \textbf{a}, Schematic of organoid differentiating protocol and human embryo experiment 
    \textbf{b-c}, UMAP based on gene expression profiles of organoid (\textbf{b}) and embryo
    (\textbf{c}) cells colored by cell type identify. 
    (Continues next page)
  }
\end{figure}
\afterpage{\clearpage}

\begin{figure}[b!]
  \ContinuedFloat
  \caption*{\rule{\linewidth}{0.4pt}}
  \caption{
    (Continued)
    \textbf{d-f}, UMAP based on gene expression profiles of organoid (top) and embryo (bottom) cells types showing sub-clusters of neural progenitors (\textbf{d}), early differentiating neurons (\textbf{e}) and neural crest (\textbf{f}) along with expression of indicated genes in red-green-blue color scale.
    \textbf{g}, UMAP based on gene expression profiles of organoid (top; UMAP axis 1 and 2 shown) and embryo (bottom; UMAP
    axis 2 and 3 shown) for neural progenitor subcluster along with the expression of indicated genes in red-green-blue
    color scale. 
    \textbf{h}, UMAP based on gene expression profiles in organoid (top left) and embryo (top right) colored by
    cell cycle phase identity and quantification of percentage of each cell cycle phase stratified by cell
    type in the organoid (bottom left) and embryo (bottom right). 
    \textbf{i}, Average gene expression level in $log_{2}(CPM)$ (dot color) and fraction of cells expressing the gene (dot size) per cell state, defined by topic modeling, for organoid
    cells (top) and embryo cells (middle) and motif enrichment level as normalized enrichment score (NES; dot color) and
    number of regions enriched for the motif (dot size) per topic (x-axis) and motif (y-axis; bottom). PVM/MIC: perivascular
    microphage/microglia.
    Color scales for RGB scatter plots are as follows (in Log(CPM)). 
    For organoid: 
    Minimum is 0 and maximum is 1.5 for all genes except:
    IRX3, BAHRL1: [0, 1.0], 
    FOXG1, GAD1, UNXC, GATA2, SOX10: [0, 0.8] and 
    ZEB2 [0, 3.0], 
    For embryo: 
    Minimum is 0 and maximum is 1.5 for all genes except:
    FOXG1, GAD1, UNCX, GATA2, SOX10 [0, 0.8], 
    ISL1 [0, 1.0], 
    ZEB2 [0, 3.0] and
    TWIST1 [0, 2.0].
  }
\end{figure}

\subsection*{
  DeepNeuralTube: sequence-to-function models that reveal the enhancer code underlying early neural and neural crest 
  development
}

To gain insight into the sequence characteristics underlying cell type-specific chromatin accessibility of 
neural progenitor, neural crest and neural cell type-specific enhancers, 
we trained two sequence-to-function model convolutional neural networks:
"$DeepNeuralTube^o$" and "$DeepNeuralTube^e$", on the organoid and embryo atlases respectively. 
The models use the DNA sequence of genomic regions as input and classify regions based on their cell type- or state-specific chromatin accessibility (in a multi-label multi-class fashion; \textbf{Fig. 2a}). 
The classes of the models are defined as the topics from topic modeling. 
To get a fine-grained definition of cell states, we performed topic modeling separately for the subset of neural progenitors, neural crest (incl. facial mesenchyme and peripheral neurons) and early differentiating neurons for both organoid (to train $DeepNeuralTube^o$) and embryo (to train $DeepNeuralTube^e$).\par

We tested the performance of $DeepNeuralTube^o$ and $DeepNeuralTube^e$ to predict cell type-specific enhancer activity 
using data from the VISTA enhancer browser\cite{viselVISTAEnhancerBrowsera2007, kosickiVISTAEnhancerBrowser2025}.
Enhancers active in the neural tube (n=349) could be classified with an area under the receiver-operator curve
(auROC) of 0.75 and 0.80 using resp. $DeepNeuralTube^o$ and $DeepNeuralTube^e$ (\textbf{Fig. 2b}; and
area under the precision-recall curve (auPR) of 0.21 resp. 0.29 (\textbf{Supplementary Fig. S2a-b})). 
Enhancers active in the facial mesenchyme (n=143) could be classified with an auROC of 0.78 and 0.79 
using resp. $DeepNeuralTube^o$ and $DeepNeuralTube^e$ (\textbf{Fig. 2c}; 
and auPR of 0.17 for both models (\textbf{Supplementary Fig. S2a-b})). 
Of note, a fraction of neural tube enhancers (\SI{24}{\percent}) are predicted to be specific to pre-migratory neural crest by
both models. Given that the VISTA enhancers are annotated based on localization of enhancer activity only, and given
that it is difficult to visually distinguish between the neural tube and pre-migratory neural crest, this result 
is not surprising.\par

Next, we assessed the explainability of the models. Concretely we tested if the model learned known
TF binding motifs as features. To this end, we calculated pairwise similarities\cite{
  guptaQuantifyingSimilarityMotifs2007} between \textit{de novo} motifs learned by the model, 
and identified using TF-MoDISco\cite{shrikumarTechnicalNoteTranscription2018}, and known motifs that are enriched in cell type specific regions.
Clustering this pairwise similarity matrix resulted in 56 clusters (\textbf{Fig. 2d}). 
All clusters, except cluster 47 (organoid specific CG motif), contain known motifs enriched in regions specific to both 
organoid and embryo cell types (\textbf{Fig. 2e} and \textbf{Supplementary Table S2}).
More importantly, most clusters (\SI{70}{\percent}) contain at least one motif learned \textit{de novo} by the organoid and/or embryo model (\textbf{Fig. 2f-g} and \textbf{Supplementary Table S2}). 
Thus, $DeepNeuralTube^o$ and $DeepNeuralTube^e$ use known-motifs as features to classify regions based
on cell type-specific chromatin accessibility.\par

As an illustrative example, we investigated an enhancer located in one of the introns of \textit{GLI3} (hs111.0; \textbf{Supplementary Fig. S2c}). 
This region has neural tube specific activity in E11.5 mouse embryos (\textbf{Fig. 2h}). 
Both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ explain this region similarly (Pearson correlation of 0.79) 
and highlight the importance of multiple SOX TFBS (\textbf{Fig. 2h}). 
A version of this enhancer where 5\% of the nucleotides are changed (hs111.1) does not have neural tube specific enhancer activity and both models predict that this loss of activity is due to the loss of a single SOX TFBS (\textbf{Fig. 2h}).\par

@@

\subsection*{
  The enhancer code underlying dorso-ventral patterning of neural progenitors
}

Focusing on neural progenitors, we identified topics representing cells and regions of different dorso-ventral
identities in both the organoid and the embryo data (\textbf{Fig. 3a}). 
From ventral to dorsal this is topic 36, 38, 33, 54 and 48 for the organoids and
topic 34, 38, 79, 88 and 58 for the embryo. 
As an example we first investigated two enhancers of \textit{SHH} that are strongly conserved and specific to the 
floorplate: SFPE1 and SFPE2\cite{jeongDistinctRegulatorsShh2003}. 
\textit{SHH} is most strongly expressed in the most ventral topic (33 and 34 resp. for the organoids and the embryo; \textbf{Fig. 3b}).
SFPE1 is located upstream of \textit{SHH} and SFPE2 in its second intron. Both enhancers are indeed most accessible in the cells belonging to the ventral topic (\textbf{Fig. 3c}). 
Explaining the accessibility of both enhancers using DeepNeuralTube reveals two FOX (one partial) and two RFX (one partial) binding sites in SFPE1 and two FOX, a ZEB and a TEAD binding site in SFPE2 (\textbf{Fig. 3d}).\par

To investigate the enhancer code underlying dorso-ventral patterning of neural progenitors further, we clustered 
and aligned \textit{de novo} motifs, identified in dorso-ventral regions, based on pairwise motif similarities. 
Next, we calculated the average contribution for all identified motif instances per cluster and for each topic 
resulting in a dorso-ventral enhancer code-table (\textbf{Fig. 3e}).
Both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ found similar patterns underlying the dorso-ventral states and the 
importance of the instances of these patterns is highly correlated across both models (0.75 average Pearson correlation; \textbf{Fig. 3e}). 
For the most ventral state (organoid topic 33 and embryo topic 34) a combination of FOX, TEAD and RFX instances 
are important (\textbf{Fig. 3e-f}) and the importance of these instances correlates with the expression of respectively \textit{FOXA1/A2/P2/P4}, \textit{TEAD1} and \textit{RFX3/4} in both the organoid and embryo (\textbf{Fig. 3g} and Supplementary Fig. S3). 
In subsequent more dorsal neural progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-1/2} expression (\textbf{(Fig. 3e- g)}); 
PAX instances, correlating with \textit{PAX3/6} expression (\textbf{Fig. 3e-g} and Supplementary Fig. S3) and 
ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig. 3e-g} and Supplementary Fig. S3). 
For all dorso-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), 
SOX instances are important correlating with the expression of \textit{SOX2/3} (\textbf{Fig. 3e-g} and Supplementary Fig. S3). 
Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig. 3e} and Supplementary Fig. S3). 
Finally, we calculated the Jaccard index of genomic regions based on the presence of instances of each identified pattern to investigate which patterns often co-occur in the same regions (\textbf{Fig. 3h}). 
FOX, TEAD and RFX instances often co-occur in the same genomic region while instances of other patterns tend to be present in separate regions (\textbf{Fig. 3h}).\par
